---
title: "Data visualisation"
---

## Introduction

Before we start with our day on Data Visualisation, create a new R script file
named `data-visualisation.R` in the folder `R/`. This will be the file
containing the R code related to Data Visualisation.

In this tutorial you'll learn:

- How to map variables in tidy data sets to visual attributes of graphics, called aesthetics in `{ggplot2}`.
- How to make different plot types by using different "geometries", called `geoms` in `{ggplot2}`.
- How to tweak the style of your plots.
- How to export your plots to files (JPEG, PNG, PDF or SVG).
- How to make article figures by combining several plots together using the `{patchwork}` package.

## The `crabs` and `quadrats` data sets

### Exercise 3.1

 - Start by loading required packages: `{tidyverse}`, `{here}` and
 `{patchwork}`. Note that the `{tidyverse}` package already includes the
 `{ggplot2}` package.
 - Read in the two tidy data sets generated yesterday: `crabs` and `quadrats`.
 Use the function `readRDS()` to read the respective `.rds` files from the
 folder `data-tidy/`.


```{r}
#| code-fold: true
#| output: false
#| code-summary: "Solution to Exercise 3.1"
library(tidyverse)
library(here)
library(patchwork)

data_tidy_path <- here::here("data-tidy")
quadrats <- readRDS(file = file.path(data_tidy_path, "quadrats.rds"))
crabs <- readRDS(file = file.path(data_tidy_path, "crabs.rds"))
```

## Mapping variables to aesthetics (aes)

The basic recipe for creating a `{ggplot2}` plot:

1. You provide the data (typically a tidy data set).
2. You indicate how to map variables to aesthetics.
3. You indicate what geoms to use.

And ggplot does the rest for you. We will now explore these three steps
while trying to answer some basic questions about the fiddler crabs.

### Exercise 3.2
What is the most abundant crab species? Make a simple bar chart showing the
number of individuals per species.

Use `dplyr::count()` to determine the number of crabs identified in each
species. The result should look like:

```{r}
#| echo: false
count(crabs, species)
```

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.2"
nr_crabs_by_species <- count(crabs, species)
ggplot(data = nr_crabs_by_species, mapping = aes(x = species, y = n)) +
  geom_col()
```

### Explanation of solution 3.2

Note how the categorical variable `species` has been mapped onto the x-axis, and
the number of crabs `n` to the y-axis.

To perform the mapping we used the function `aes()` and then passed the result
to the `mapping` parameter of the function `ggplot()`. We can pry on what
kind of information `aes()` is generating by evaluating it here in isolation:

```{r}
aes(x = species, y = n)
```

Also note that the observational unit of the data set `nr_crabs_by_species` is
now the species, not an individual crab as we had in the case of the `crabs`
data set. So we have three observations (each species), and each one corresponds
to one bar in the plot.

### Exercise 3.3

What are the abundances of each sex?

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.3"
nr_crabs_by_sex <- count(crabs, sex)
ggplot(data = nr_crabs_by_sex, mapping = aes(x = sex, y = n)) +
  geom_col()
```

Okay, so you've re-mapped the x-axis to the `sex` variable in the
`nr_crabs_by_sex` data set.

### Exercise 3.4

Can you visually show the crab abundance by species and sex in the same plot? If
you look at the plot, what does it tell you about how easy it might be to
distinguish female and male crabs across species in the field?

> Tip: pay attention to the number of `NA` values in each species.

To determine the number of crabs by species and sex, just use `dplyr::count()`
to count the number of crabs and ask for the combination of values for `species`
and `sex`:

```{r}
dplyr::count(crabs, species, sex)
```

And now we can make a new bar plot with the following mappings:

- `species` maps to bars' x-position.
- `sex` maps to fill colour of the bars (the `fill` aesthetic).
- `n` maps to number of crabs, as usual.

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.4"
nr_crabs_by_sp_sex <- count(crabs, species, sex)
ggplot(data = nr_crabs_by_sp_sex, mapping = aes(x = species, y = n, fill = sex)) +
  geom_col()
```

Nice! We got a stacked bar plot. `sex` is now mapped on the fill colour in the
bar plot. Here are variations (V1-V5) on how we could tweak the
width and positioning of the bars regarding the `sex` variable.

::: {.panel-tabset}

### V1

```{r}
nr_crabs_by_sp_sex %>%
  ggplot(mapping = aes(x = species, y = n, fill = sex)) +
  geom_col(position = "dodge")
```

### V2

```{r}
nr_crabs_by_sp_sex %>%
  ggplot(mapping = aes(x = species, y = n, fill = sex)) +
  geom_col(position = "dodge2")
```

### V3

```{r}
nr_crabs_by_sp_sex %>%
  ggplot(mapping = aes(x = species, y = n, fill = sex)) +
  geom_col(position = position_dodge2(preserve = "single"))
```

### V4

```{r}
nr_crabs_by_sp_sex %>%
  tidyr::complete(species, sex) %>%
  ggplot(mapping = aes(x = species, y = n, fill = sex)) +
  geom_col(position = position_dodge2(preserve = "single"))
```

### V5

```{r}
nr_crabs_by_sp_sex %>%
  tidyr::complete(species, sex, fill = list(n = 0)) %>%
  ggplot(mapping = aes(x = species, y = n, fill = sex)) +
  geom_col(position = position_dodge2(preserve = "single"))
```

:::

### Exercise 3.5

The latest plots were showing three variables:

- `species` on the x-axis
- `sex` as fill colour
- `n` on the y-axis

Let us see how far we may go by mapping more variables to other aesthetics.
Besides the variables `species` and `sex`, `stage` (crab developmental stage) is
also an important demographics' variable. Let us add that too to the mix. There is an
aesthetic called `colour` that adjusts the colour of the bar outline.

So, try to make a new plot where, in addition, the `colour` aesthetic is mapped
to `stage`.

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.5"
count(crabs, species, sex, stage) %>%
  ggplot(mapping = aes(x = species, y = n, fill = sex, colour = stage)) +
  geom_col(position = "dodge2")
```
What an ugly plot! Perhaps trying to fit too many variables into one single
plot may not be such a great idea.

### Exercise 3.6

There is one powerful technique to improve readability of your plots. If one
plot is not enough, make several! This can be achieved with
faceting. Faceting is the process of creating several related plots, each
in their own facet (panel).

Here is an example of three facets, one for each species:

```{r}
count(crabs, species, stage) %>%
  ggplot(mapping = aes(x = stage, y = n)) +
  geom_col(position = "dodge2") +
  facet_wrap(vars(species))
```

Now, try to make the ugly plot again, but now using only three aesthetic
mappings while leaving the species to be varied across facets:

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.6"
ggplot(data = count(crabs, species, sex, stage), mapping = aes(x = sex, y = n, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_wrap(vars(species))
```
Hooray! Four variables shown in not such a complicated way. But we can do
better...

### Exercise 3.7

Let us draw five variables: `species`, `sex`, `stage`, `season` and `n`.
We can use faceting in 2D: rows and columns. Take a look at the solution
below.

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.7"
count(crabs, species, sex, stage, season) %>%
ggplot(mapping = aes(x = sex, y = n, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(rows = vars(season), cols = vars(species))
```

Based on this plot can you tell:

- What developmental stages are more abundant? Sexual maturation of the fiddler
crabs occurs from `pre_puberty` to `sub_adult`.  It seems there is quite a
significant different between the abundance of the stages prior to sexual
maturation and posterior to that. Can you perhaps put forward an explanation
by speculating on the relative duration of each stage.

- Also, does the crab abundance also change between summer and winter? How so?

Note that, by default, when you do faceting, `{ggplot2}` will automatically
pick the range (or order) of the values in the x-axis and y-axis such that
all plots are readily comparable. Although this is generally a desirable
feature, it might also make answering specific questions more difficult.
For example, the relative abundances across stages in the winter are difficult
to appreciate because the overall abundance in the winter is so reduced. It
would be nice if we could change the y-axis of the winter panels to show
the full range.

### Exercise 3.8

Look up the documentation of the `facet_grid()` and find the parameter taken
by this function that allows you to free the y-axis to be expressed in its
full range.

Is the pattern of crab abundance across `stage` more extreme in the summer or in
winter?

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.8"

# The `facet_grid()` parameter that allows you to free the y-axis is: `scales`.
# 
# Allowing the y-scale to be different between summer and winter allows us
# to better appreciate differences across stages in the winter season.
ggplot(data = count(crabs, species, sex, stage, season), mapping = aes(x = sex, y = n, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single"), na.rm = TRUE) +
  scale_x_discrete(na.translate = FALSE) +
  facet_grid(rows = vars(season), cols = vars(species), scales = "free_y")
```

It seems that in the winter season we find mostly adult crabs, followed by a
small percentage of sub-adult crabs, and we are very unlikely to find juveniles
or pre-puberty crabs.

Perhaps it would be nice to show these percentages instead. The absolute abundances
are, after all, influenced by the sampling effort by the marine biologist, and 
the sampling area of the quadrats.

### Exercise 3.9

Use a combination of `dplyr::count()`, `dplyr::group_by()` and `dplyr::mutate()`
to create a new variable (i.e. column) for the percentage of crabs within each
panel (i.e. species-season combination). Name this new column `pct`
(abbreviation of percentage).

```{r}
#| code-fold: true
#| code-summary: "Solution to Exercise 3.9"

# To further distinguish the distribution across stages we might want to normalise
# with respect to the species-season combination and show a percentage.
count(crabs, species, sex, stage, season) %>%
  group_by(species, season) %>%
  mutate(pct = 100 * n / sum(n, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = sex, y = pct, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single"), na.rm = TRUE) +
  scale_x_discrete(na.translate = FALSE) +
  facet_grid(rows = vars(season), cols = vars(species), scales = "free_y")
```

Take a look at the variations below (V1--V3) and pay attention to the small
code changes and their effects on the style of the plot.

::: {.panel-tabset}

### V1

```{r}
# Polish the scales' labels.
count(crabs, species, sex, stage, season) %>%
  group_by(species, season) %>%
  mutate(frac = n / sum(n, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = sex, y = frac, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single"), na.rm = TRUE) +
  scale_x_discrete(na.translate = FALSE, labels = c("Female", "Male")) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_ordinal(labels = c("Juvenile", "Pre-puberty", "Sub-adult", "Adult")) +
  labs(x = 'Crab sex', y = "Population percentage", fill = "Dev. stage") +
  facet_grid(rows = vars(season), cols = vars(species), scales = "free_y")
```

### V2

```{r}
count(crabs, species, sex, stage, season) %>%
  group_by(species, season) %>%
  mutate(frac = n / sum(n, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = sex, y = frac, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single"), na.rm = TRUE) +
  scale_x_discrete(na.translate = FALSE, labels = c("Female", "Male")) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_ordinal(limits = rev, labels = rev(c("Juvenile", "Pre-puberty", "Sub-adult", "Adult"))) +
  labs(x = 'Crab sex', y = "Population percentage", fill = "Dev. stage") +
  facet_grid(rows = vars(season), cols = vars(species), scales = "free_y")
```

### V3

```{r}
count(crabs, species, sex, stage, season) %>%
  group_by(species, season) %>%
  mutate(frac = n / sum(n, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = sex, y = frac, fill = stage)) +
  geom_col(position = position_dodge2(preserve = "single"), na.rm = TRUE) +
  scale_x_discrete(na.translate = FALSE, labels = c("Female", "Male")) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_hue(labels = c("Juvenile", "Pre-puberty", "Sub-adult", "Adult")) +
  labs(x = 'Crab sex', y = "Population percentage", fill = "Dev. stage") +
  facet_grid(rows = vars(season), cols = vars(species), scales = "free_y")
```

:::
